### 2020-06-20 article.tex
Avoid copy to GPU in PetscLogStage

% FundRef data to be entered by SIAM
%<funding-group specific-use="FundRef">
%<award-group>
%<funding-source>
%<named-content content-type="funder-name"> 
%</named-content> 
%<named-content content-type="funder-identifier"> 
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>
\newcommand\col{:}
\newcommand{\pk}[1]{{\color{gray!90!black} #1}}
\usepackage{stmaryrd}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\pgfplotsset{colormap={paraview}{rgb(0cm)=(0.278431,0.278431,0.858824) rgb(0.1428571429cm)=(0,0,0.360784) rgb(0.2857142857cm)=(0,1,1) rgb(0.4285714286cm)=(0,0.501961,0) rgb(0.5714285714cm)=(1,1,0) rgb(0.7142857143cm)=(1,0.380392,0) rgb(0.8571428571cm)=(0.419608,0,0) rgb(1cm)=(0.878431,0.301961,0.301961)}}
\usepackage{fancyvrb}
\usepackage{subfig}
\usepackage{upquote,textcomp}
\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\definecolor{last}{RGB}{44,123,182}
\definecolor{blueaccent}{RGB}{215,25,28}
\definecolor{myyellow}{RGB}{102,102,17}
\definecolor{myred}{RGB}{197,65,0}
\definecolor{myblue}{RGB}{0,0,235}
\definecolor{myaqua}{RGB}{9,156,156}
\definecolor{mygreen}{RGB}{63,126,28}
\definecolor{paraviewred}{rgb}{0.278431,0.278431,0.858824}
\definecolor{paraviewblue}{rgb}{0.878431,0.301961,0.301961}
\definecolor{darkgray}{RGB}{253,174,97}
\definecolor{mediumgray}{RGB}{255,255,191}
\usepackage[outline]{contour}
\usepackage{tikz}
\usepackage[binary-units=true]{siunitx}
\usetikzlibrary{decorations.text,arrows.meta,calc,plotmarks,shadows,spy,fixedpointarithmetic,patterns,intersections,arrows.meta,shapes,decorations.text,decorations.pathmorphing,backgrounds,fit,positioning,shapes.symbols,chains,3d,calc,external}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}
\makeatother
\usepackage{amsmath,etoolbox}
\AtBeginEnvironment{pmatrix}{\setlength{\arraycolsep}{2pt}}
\pgfmathsetmacro{\dof}{4}
\pgfmathsetmacro{\overlap}{1}
\begin{document}

@@

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced together in order to offer
    to end-users adaptive Krylov and overlapping Schwarz methods.  The
    flexibility of the implementation %, which for example may deal with various matrix formats, 
    are showcased and explained.  Diverse examples are
    displayed, ranging from fracture mechanics to linear stability analysis, and
    the sets of options are thoroughly documented, so that they may be readily
    tried on other applications and reproduce results.
\end{abstract}

@@

\section{Introduction}
Implementing and maintaining robust and scalable applications is highly relevant
in the field of computational science and engineering. Some tools exist to help
end-users achieve this goal. One the one hand, PETSc \cite{balay1997cient}, the
Portable and Extensible Toolkit for Scientific computation, is a
well-established software from the community. It may be used to efficiently
discretize partial differential equations, or solve algebraic linear or
nonlinear systems of equations. It comes with many built-in features such as
advanced preconditioning or tailored matrix formats. Also, it can % be used to compile and 
interact with third-party libraries, such as
\emph{hypre}~\cite{falgout2002hypre} for linear solvers, TetGen~\cite{si2013tet}
for mesh generation, and so on. These appealing features convinced computational
scientists to use PETSc as one of the discretization and/or algebraic backend in
many different higher-level projects, such as finite element or finite volume
packages. On the other hand, HPDDM is a much smaller project purely focused on
domain decomposition preconditioners~\cite{jolivet2013scalable} % , a small subclass of preconditioners available in PETSc as well, 
and advanced iterative
methods~\cite{jolivet2016block}. Such methods deal efficiently with linear
systems when multiple right-hand sides are available simultaneously, and are
then referred to as block iterative methods, or when there is a recurrence of
varying coefficient matrices and right-hand sides, and are then referred to as
recycling iterative methods.
% Htool ?
In this paper, the integration of both aspects of HPDDM, its advanced iterative
methods and its domain decomposition preconditioners, into PETSc is explained.
The design principles of the interface are showcased, with % alongside
minimalist and easy-to-run examples. The paper is divided in two main parts:
\cref{sec:KSPHPDDM} which introduces \pk{KSPHPDDM}, and \cref{sec:PCHPDDM} which
introduces \pk{PCHPDDM}. Eventually, concluding remarks may be found
\cref{sec:conclusion}. All PETSc keywords are highlighted in \pk{gray}, these
are public and documented at \url{https://www.mcs.anl.gov/petsc/petsc-master/docs}.
\section{KSPHPDDM}\label{sec:KSPHPDDM}
  \subsection{Related work}
Krylov subspace methods are widely used in numerical linear algebra for solving
systems of equations. Some of their strengths are memory efficiency and the fact
that they are mostly format agnostic. Indeed, they mainly rely on matrix--vector
operations such as multiplications or transpose multiplications. For
challenging high-dimensional systems, finding appropriate preconditioners is
mandatory to make these methods robust. %, but this goes beyond the scope of this section.
In PETSc, Krylov methods belong to the \pk{KSP} base class. Some of the
most used types are \pk{KSPGMRES}, \pk{KSPCG}, or \pk{KSPBCGS}, which respectively implement
the generalized minimal residual method~\cite{saad1986gmres}~(GMRES), conjugate
gradient~\cite{hestenes1952methods}~(CG), or biconjugate gradient stabilized method~\cite{van1992bi}~(BCGS). % Some of these methods are also
% available in other linear algebra backends such as~\emph{hypre} or DUNE~\cite{bastian2010generic}.

@@

% Recycling put aside, 
Another important aspect of Krylov methods is their ability
to deal with multiple right-hand sides simultaneously. Block Krylov methods~\cite{gutknecht2006block} are
designed for solving linear systems $A X = B$, where
$X$ and $B$ are tall-and-skinny matrices with $k \geq 1$ columns. These methods are
less frequently implemented in general purpose libraries because they require
somehow more involved kernels such as matrix--matrix multiplication, instead of
matrix--vector. Still, systems with multiple right-hand sides are ubiquitous,
e.g., in tomography~\cite{tournier2016micro} or data
analytics~\cite{KALANTZIS2018136}. % Of course, it is possible to cast a system with multiple right-hand sides into a sequence of systems with the same coefficient matrix but changing right-hand sides $A x_i = b_i, i = 1,2,\ldots,k$. However, 
Block Krylov methods, while having higher arithmetic
intensities, generate larger subspaces and typically converge in fewer iterates.
There is none implemented in PETSc. It is nevertheless possible to solve
systems with multiple right-hand sides using a specialized type of matrices,
but with a very limited choice of preconditioners, cf.~\cref{sec:matkaij}.
In the context of optimization, it is also
frequent to solve a sequence of linear systems $A_i X_i = B_i, i = 1, 2,
\ldots$

@@

flag -{}-download-hpddm. It is then possible to select the corresponding \pk{KSP}
using the command line option -ksp\_type hpddm, or the routine \pk{KSPSetType}(ksp,

@@

their ``standard'' counterpart. These methods may be selected using the additional
options -ksp\_hpddm\_type (gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly) and
-ksp\_hpddm\_variant (left$|$right$|$flexible) to select whether preconditioning

@@

this option through the more common PETSc option -ksp\_pc\_side (left$|$right). 

@@

to compute $X = P_BY$ if the preconditioner is of one of the following type:

@@

    \item block Jacobi with a single subdomain per process and exact or
        incomplete subdomain solvers, \pk{PCBJACOBI};
    \item overlapping Schwarz methods with a single subdomain per process and

@@

and call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$). With the two first sets of
preconditioners, $\tilde{Y} = Y$ and $\tilde{X} = X$. With block Jacobi,
$\tilde{Y}$, resp. $\tilde{X}$, represents only the part of the matrix $Y$,
resp. $X$, local to a given process. With overlapping Schwarz methods, there are
furthermore overlapping ``ghost'' unknowns, interested readers are referred to
monographs on practical aspects of domain decomposition
methods~\cite{smith2004domain,dolean2014ddm}. For all other types of
preconditioners, HPDDM will loop over the columns of $Y$ and apply $P_B$
successively to get all columns of $X$.

The rest of the operations are then performed directly inside HPDDM. For
example, in all Krylov methods using the Arnoldi process, it is mandatory to orthonormalize
candidate basis vectors. It would be impractical to try to reuse PETSc routines
like \pk{KSPGMRESModifiedGramSchmidtOrthogonalization} for two reasons: 1) it does
not handle the orthonormalization of multiple vectors at once which is needed in
pseudo-block methods 2) HPDDM may be used without PETSc as a stand-alone
library. To give some more technical details to curious readers, in HPDDM, $QR$
factorizations of tall-and-skinny dense matrices, which are obviously needed for
block Krylov methods but also for recycling Krylov methods are computed using
the CholQR algorithm~\cite{stathopoulos2002block}, or the block modified or classical
Gram--Schmidt method. This may be adjusted with the option -ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs).
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.

To monitor convergence of a \pk{KSP} object, calls to the PETSc \pk{KSPMonitor} routine
were added inside HPDDM Krylov methods. 

@@

subsequent solve, this information is reused appropriately. Two new routines were

@@

multiple right-hand sides instead. The standard function \pk{KSPSolve} takes as
input argument a \pk{KSP} and two \pk{Vec}. A new function
\pk{KSPMatSolve} was added in PETSc 3.14.0, which takes as input argument

@@

Otherwise, it is possible to use any \pk{KSPHPDDM} (pseudo-)block Krylov methods.

@@

    \item in MATLAB, unpreconditioned GCRODR(40, 20), $ICC(0)$ left-preconditioned

@@

        legend style={font=\small,at={(1.02,0.25)}, legend cell align=left,
        anchor=south west}]

@@

In SLEPc, these systems may be parameterized using the -st\_ prefix, e.g.,
-st\_pc\_type~lu. If they are solved iteratively, it is possible to

@@

Indeed, through the 46 inner solves, these numbers are roughly decreased by 25. In terms of runtime, \SI[round-mode          = places,round-precision=1]{1.5098e+03}{\sec} (resp.~\SI[round-mode          = places,round-precision=1]{1.0524e+03}{\sec}) is spent in \pk{EPSSolve} when using \pk{KSPFGMRES} (resp.~\pk{KSPHPDDM}). These timings, and all which follow, have been obtained on {Ir\`ene}, a system composed of \pgfmathprintnumber[assume math mode=true]{1656} nodes with two 24-core Intel Xeon Platinum 8168 clocked at \SI{2.7}{\giga\hertz}. 

@@

@@

        legend style={font=\small,at={(0.5,0.24)}, legend cell align=left,
        anchor=south west}]

@@

        \caption{\# of inner iterations for Krylov--Schur systems~\cref{eq:ks} with~\ref{pgfplots:m0} or without~\ref{pgfplots:m1} recycling.\label{fig:comparison-slepc}}

@@

1      2.0126e-01   1.4696e+08 8.0098e-02   1.4696e+08   9.3388e-02   1.4696e+08     8.9584e-01   1.3535e+09    6.2428e-01   1.3638e+09     1.2072e+00   1.3535e+09   4.0734e+00   5.4450e+09    2.2476e+00   5.4708e+09    5.0428e+00   5.4450e+09   4.4800e-02   1.5211e+08   3.7933e-01   1.3690e+09   1.5043e+00   5.4759e+09   6.6857e-02  7.4649e-01   3.1128e+00   4.7672e-01   6.2903e-01  2.2271e+00   4.7675e-01   6.2990e-01    2.2273e+00    7.8399e-03    8.6487e-02    3.5536e-01   5.2814e-02 6.6225e-02   2.4584e-01   4.5115e-02   6.6404e-02   2.4489e-01

@@

    \subsubsection{Performance of primitives for block Krylov methods}

@@

To keep this study succinct, only sequential sparse matrices will be evaluated next, which can help draw conclusions for the intraprocess performance of parallel sparse matrices. The Seq substring is now dropped, since in this case, \pk{MatSeqAIJ} is equivalent to \pk{MatAIJ} and so on.

The Poisson equation on a cube is discretized with $\mathbb{P}_1$ finite elements in double precision. This yields a matrix $A$ of dimension one million. Though the numerical values of $A$ are of no interest here, the sparsity pattern is rather common and frequently encountered, e.g., when discretizing partial differential equations. Then, a random symmetric dense matrix $T$ of dimension $b$ is generated, for $b \in \{ 1,3,6 \}$, and the following matrix is evaluated $\mathbb{A} = A \otimes T$. The matrix $\mathbb{A}$ is then assembled into all three formats described above. For block matrices, a block size of $b$ is used. The performance of the \pk{MatProductNumeric} operation is evaluated for tall-and-skinny dense matrices with a varying number of columns $N \in \{1,2,8,16,32,64\}$. For $N=1$, the \pk{MatMult} operation for matrix--vector multiplication is used instead. All results are gathered in~\cref{fig:perfAX}. They can be reproduced using the following mini-app: \url{http://jolivet.perso.enseeiht.fr/MatProduct.cpp} and any input \pk{MatAIJ} stored in binary format, here using the default name binaryoutput.

@@

  \pgfmathparse{1.3535e+09/8.9584e-01/1000000000.0}\edef\storeresultII{\pgfmathresult}
  \pgfmathparse{5.4450e+09/4.0734e+00/1000000000.0}\edef\storeresultIII{\pgfmathresult}

@@

  % \hspace*{-1.5cm}

@@

        ytick = {1,2,3,4},

@@

          \addplot+[draw=black,fill, area legend,postaction={pattern=crosshatch dots}] table [x expr=\thisrow{N}>7?\thisrow{N}/2:\thisrow{N},y expr=\thisrow{aij-3-F}/\thisrow{aij-3-T}/(\storeresultII*1000000000.0)]{\loadedMatProduct};\label{pgfplots:aij3}
          \addplot+[nodes near coords style={xshift=0.5cm,yshift=-0.1cm,above}, nodes near coords*={\rotatebox{45}{\tiny\color{black}{\pgfmathfloattofixed{\pgfplotspointmeta}\pgfmathparse{int(round(\pgfmathresult))}\ifnum\pgfmathresult=1\pgfmathprintnumber\storeresultII~GFLOP/s\fi}}},

@@

           % \addlegendentry{\pk{MatAIJ}}
           % \addlegendentry{\pk{MatBAIJ}}
           % \addlegendentry{\pk{MatSBAIJ}}
           % \addlegendentry{\pk{MatAIJMKL}}
           % \addlegendentry{\pk{MatBAIJMKL}}
           % \addlegendentry{MatSBAIJMKL}

@@

        ytick = {1,2,3,4},

@@

          \addplot+[draw=black,fill, area legend,postaction={pattern=crosshatch dots}] table [x expr=\thisrow{N}>7?\thisrow{N}/2:\thisrow{N},y expr=\thisrow{aij-10-F}/\thisrow{aij-10-T}/(\storeresultIII*1000000000.0)]{\loadedMatProduct};\label{pgfplots:aij10}
          \addplot+[nodes near coords style={xshift=0.5cm,yshift=-0.1cm,above}, nodes near coords*={\rotatebox{45}{\tiny\color{black}{\pgfmathfloattofixed{\pgfplotspointmeta}\pgfmathparse{int(round(\pgfmathresult))}\ifnum\pgfmathresult=1\contour{white}{\pgfmathprintnumber\storeresultIII}~GFLOP/s\fi}}},draw=black,fill, area legend,postaction={pattern=crosshatch dots}] table [x expr=\thisrow{N}>7?\thisrow{N}/2:\thisrow{N},y expr=\thisrow{baij-10-F}/\thisrow{baij-10-T}/(\storeresultIII*1000000000.0)]{\loadedMatProduct};\label{pgfplots:baij10}

@@

   \caption{Performance of the matrix--matrix multiplication as implemented in PETSc for three sparse matrix types. As baseline values with $N=1$, for $b=1$, performance of \pk{MatAIJ} matrix--vector multiplication is used~\ref{pgfplots:aij1}, while for $b=\{3,6\}$, it is the performance of \pk{MatBAIJ} matrix--vector multiplication~\ref{pgfplots:baij3}.\label{fig:perfAX}}

@@

>>>>

% FundRef data to be entered by SIAM
%<funding-group specific-use="FundRef">
%<award-group>
%<funding-source>
%<named-content content-type="funder-name"> 
%</named-content> 
%<named-content content-type="funder-identifier"> 
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>
\newcommand\col{:}
\newcommand{\pk}[1]{\texttt{#1}}
\usepackage{stmaryrd}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\pgfplotsset{colormap={paraview}{rgb(0cm)=(0.278431,0.278431,0.858824) rgb(0.1428571429cm)=(0,0,0.360784) rgb(0.2857142857cm)=(0,1,1) rgb(0.4285714286cm)=(0,0.501961,0) rgb(0.5714285714cm)=(1,1,0) rgb(0.7142857143cm)=(1,0.380392,0) rgb(0.8571428571cm)=(0.419608,0,0) rgb(1cm)=(0.878431,0.301961,0.301961)}}
\usepackage{fancyvrb}
\usepackage{subfig}
\usepackage{upquote,textcomp}
\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\definecolor{last}{RGB}{44,123,182}
\definecolor{blueaccent}{RGB}{215,25,28}
\definecolor{myyellow}{RGB}{102,102,17}
\definecolor{myred}{RGB}{197,65,0}
\definecolor{myblue}{RGB}{0,0,235}
\definecolor{myaqua}{RGB}{9,156,156}
\definecolor{mygreen}{RGB}{63,126,28}
\definecolor{paraviewred}{rgb}{0.278431,0.278431,0.858824}
\definecolor{paraviewblue}{rgb}{0.878431,0.301961,0.301961}
\definecolor{darkgray}{RGB}{253,174,97}
\definecolor{mediumgray}{RGB}{255,255,191}
\usepackage[outline]{contour}
\usepackage{tikz}
\usepackage[binary-units=true]{siunitx}
\usetikzlibrary{decorations.text,arrows.meta,calc,plotmarks,shadows,spy,fixedpointarithmetic,patterns,intersections,arrows.meta,shapes,decorations.text,decorations.pathmorphing,backgrounds,fit,positioning,shapes.symbols,chains,3d,calc,external}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}
\makeatother
\usepackage{amsmath,etoolbox}
\AtBeginEnvironment{pmatrix}{\setlength{\arraycolsep}{2pt}}
\pgfmathsetmacro{\dof}{4}
\pgfmathsetmacro{\overlap}{1}
\begin{document}

@@

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced together in order to offer
    to end-users adaptive Krylov and overlapping Schwarz methods.  The
    flexibility of the implementation %, which for example may deal with various matrix formats, 
    is showcased and explained.  Diverse examples are
    displayed, ranging from fracture mechanics to linear stability analysis, and
    the sets of options are thoroughly documented, so that they may be readily
    tried on other applications and reproduce results.
\end{abstract}

@@

\section{Introduction}
Implementing and maintaining robust and scalable applications is highly relevant
in the field of computational science and engineering. Some tools exist to help
end-users achieve this goal. One the one hand, PETSc \cite{balay1997cient}, the
Portable and Extensible Toolkit for Scientific computation, is a
well-established software from the community. It may be used to efficiently
discretize partial differential equations, or solve algebraic linear or
nonlinear systems of equations. It comes with many built-in features such as
advanced preconditioning or tailored matrix formats. Also, it can % be used to compile and 
interact with third-party libraries, such as
\emph{hypre}~\cite{falgout2002hypre} for linear solvers, TetGen~\cite{si2013tet}
for mesh generation, and so on. These appealing features convinced computational
scientists to use PETSc as one of the discretization and/or algebraic backend in
many different higher-level projects, such as finite element or finite volume
packages. On the other hand, HPDDM is a much smaller project purely focused on
domain decomposition preconditioners~\cite{jolivet2013scalable} % , a small subclass of preconditioners available in PETSc as well, 
and advanced iterative
methods~\cite{jolivet2016block}. Such methods deal efficiently with linear
systems when multiple right-hand sides are available simultaneously, and are
then referred to as block iterative methods, or when there is a recurrence of
varying coefficient matrices and right-hand sides, and are then referred to as
recycling iterative methods.
% Htool ?
In this paper, the integration of both aspects of HPDDM, its advanced iterative
methods and its domain decomposition preconditioners, into PETSc is explained.
The design principles of the interface are showcased, with % alongside
minimalist and easy-to-run examples. The paper is divided in two main parts:
\cref{sec:KSPHPDDM} which introduces \pk{KSPHPDDM}, and \cref{sec:PCHPDDM} which
introduces \pk{PCHPDDM}. Eventually, concluding remarks may be found
\cref{sec:conclusion}. All PETSc keywords and options are typeset in a \texttt{typewriter} front, these
are public and documented at \url{https://www.mcs.anl.gov/petsc/petsc-master/docs}.
\section{KSPHPDDM}\label{sec:KSPHPDDM}
  \subsection{Related work}
Krylov subspace methods are widely used in numerical linear algebra for solving
systems of equations. Some of their strengths are memory efficiency and the fact
that they are mostly format agnostic. Indeed, they mainly rely on matrix--vector
operations such as multiplications or transpose multiplications. For
challenging high-dimensional systems, finding appropriate preconditioners is
mandatory to make these methods robust. %, but this goes beyond the scope of this section.
In PETSc, Krylov methods belong to the \pk{KSP} base class. Some of the
most used types are \pk{KSPGMRES}, \pk{KSPCG}, or \pk{KSPBCGS}, which respectively implement
the generalized minimal residual method~\cite{saad1986gmres}~(GMRES), conjugate
gradient~\cite{hestenes1952methods}~(CG), or biconjugate gradient stabilized method~\cite{van1992bi}~(BCGS). % Some of these methods are also
% available in other linear algebra backends such as~\emph{hypre} or DUNE~\cite{bastian2010generic}.

@@

% Recycling put aside, 
Another important aspect of Krylov methods is their ability
to deal with multiple right-hand sides simultaneously. Block Krylov methods~\cite{gutknecht2006block} are
designed for solving linear systems $A X = B$, where
$X$ and $B$ are tall-and-skinny matrices with $k \geq 1$ columns. These methods are
less frequently implemented in general purpose libraries because they require
somehow more involved kernels such as matrix--matrix multiplication, instead of
matrix--vector. Still, systems with multiple right-hand sides are ubiquitous,
e.g., in tomography~\cite{tournier2016micro}, data
analytics~\cite{KALANTZIS2018136}, eigensolvers~\cite{knyazev2001toward}, geophysics~\cite{calandra2012flexible}, quantum chromodynamics~\cite{sakurai2010application}, etc. % Of course, it is possible to cast a system with multiple right-hand sides into a sequence of systems with the same coefficient matrix but changing right-hand sides $A x_i = b_i, i = 1,2,\ldots,k$. However, 
Block Krylov methods, while having higher arithmetic
intensities, generate larger subspaces and typically converge in fewer iterates.
There is none implemented in PETSc. It is nevertheless possible to solve
systems with multiple right-hand sides using a specialized type of matrices,
but with a very limited choice of preconditioners, cf.~\cref{sec:matkaij}.
In the context of optimization, it is also
frequent to solve a sequence of linear systems $A_i X_i = B_i, i = 1, 2,
\ldots$

@@

flag \pk{-{}-download-hpddm}. It is then possible to select the corresponding \pk{KSP}
using the command line option \pk{-ksp\_type hpddm}, or the routine \pk{KSPSetType}(ksp,

@@

their ``standard'' counterpart. These methods may be selected using the 
options \pk{-ksp\_hpddm\_type} \pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)} and
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether preconditioning

@@

this option through the more common PETSc option \pk{-ksp\_pc\_side (left$|$right)}. 

@@

to compute $X = P_BY$ if the preconditioner is of one of the following types:

@@

    \item block Jacobi with % a single subdomain per process and
        exact or incomplete subdomain solvers, \pk{PCBJACOBI};
    \item overlapping Schwarz methods with % a single subdomain per process and

@@

and call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$). % With the two first sets of preconditioners, $\tilde{Y} = Y$ and $\tilde{X} = X$. With block Jacobi, $\tilde{Y}$, resp. $\tilde{X}$, represents only the part of the matrix $Y$, resp. $X$, local to a given process. With overlapping Schwarz methods, there are furthermore overlapping ``ghost'' unknowns, interested readers are referred to monographs on practical aspects of domain decomposition methods~\cite{smith2004domain,dolean2014ddm}.
For all other types of preconditioners, HPDDM will loop over the columns of $Y$
and apply $P_B$ successively to get all columns of $X$.

The rest of the operations are then performed directly inside HPDDM. For
example, in all Krylov methods using the Arnoldi process, it is mandatory to orthonormalize
candidate basis vectors. % It would be impractical to try to reuse PETSc routines like \pk{KSPGMRESModifiedGramSchmidtOrthogonalization} for two reasons: 1) it does not handle the orthonormalization of multiple vectors at once which is needed in pseudo-block methods 2) HPDDM may be used without PETSc as a stand-alone library. To give some more technical details to curious readers, in HPDDM,
$QR$ factorizations of tall-and-skinny dense matrices, which are needed by
block Krylov methods and for recycling Krylov methods are computed using
the CholQR algorithm~\cite{stathopoulos2002block}, or the block modified or classical
Gram--Schmidt method. This may be adjusted with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}.
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.

Monitoring convergence with the \pk{KSPMonitor} and \pk{KSPSetConvergenceTest} callbacks is fully supported.

@@

subsequent solve, this information is reused appropriately~\cite{saad2000deflated,stathopoulos2002block,soodhalter2014krylov}. Two new routines were

@@

multiple right-hand sides instead. % The standard function \pk{KSPSolve} takes as input argument a \pk{KSP} and two \pk{Vec}.
A new function \pk{KSPMatSolve} was added in PETSc 3.14.0, which takes as input argument

@@

Otherwise, it is possible to use any \pk{KSPHPDDM} \mbox{(pseudo-)block} Krylov methods.

@@

    \item in MATLAB, unpreconditioned GCRODR(40, 20), ICC(0) left-preconditioned

@@

        legend style={font=\small,at={(1.02,0.5)},anchor=west, legend cell align=left}]

@@

In SLEPc, these systems are parameterized using the \pk{-st\_} prefix, e.g.,
\pk{-st\_pc\_type~lu}. If they are solved iteratively, it is possible to

@@

Indeed, through the 46 inner solves, these numbers are roughly decreased by 25. In terms of runtime, \SI[round-mode          = places,round-precision=1]{1.5098e+03}{\sec} (resp.~\SI[round-mode          = places,round-precision=1]{1.0524e+03}{\sec}) is spent in \pk{EPSSolve} when using \pk{KSPFGMRES} (resp.~\pk{KSPHPDDM}). These timings, and all which follow except in~\cref{sec:primitives}, have been obtained on {Ir\`ene}, a system composed of \pgfmathprintnumber[assume math mode=true]{1656} nodes with two 24-core Intel Xeon Platinum 8168 clocked at \SI{2.7}{\giga\hertz}. 

@@

    \begin{figure}[!b]

@@

        legend columns=1,enlarge x limits=0.05, height=5.5cm,
        legend style={font=\small,at={(0.98,0.02)}, legend cell align=left,
        anchor=south east}]

@@

        \caption{\# of inner iterations for systems~\cref{eq:ks} with~\ref{pgfplots:m0} or without~\ref{pgfplots:m1} recycling.\label{fig:comparison-slepc}}

@@

1      2.0126e-01   1.4696e+08 8.0098e-02   1.4696e+08   9.3388e-02   1.4696e+08     8.9584e-01   1.3535e+09    6.2428e-01   1.3638e+09     1.2072e+00   1.3535e+09   4.0734e+00   5.4450e+09    2.2476e+00   5.4708e+09    5.0428e+00   5.4450e+09   2.1248e-03   1.5211e+08   1.3669e-02   1.3690e+09   4.7833e-02   5.4759e+09   6.6857e-02  7.4649e-01   3.1128e+00   4.7672e-01   6.2903e-01  2.2271e+00   4.7675e-01   6.2990e-01    2.2273e+00    7.8399e-03    8.6487e-02    3.5536e-01   5.2814e-02 6.6225e-02   2.4584e-01   4.5115e-02   6.6404e-02   2.4489e-01

@@

    \subsubsection{Performance of primitives for block Krylov methods}\label{sec:primitives}

@@

To keep this study succinct, only nondistributed sparse matrices will be evaluated next, which can help draw conclusions for the intraprocess performance of parallel sparse matrices. The Seq substring is now dropped, since in this case, \pk{MatSeqAIJ} is equivalent to \pk{MatAIJ} and so on. The performance of the multiplication primitives of the three aforementioned types will also be compared to MKL inspector--executor sparse BLAS routines~\cite{mkl-web-page}, using a single OpenMP thread. There is an ongoing effort to better integrate these routines in PETSc respective types \pk{MatAIJMKL} and \pk{MatBAIJMKL}, but at the time of writing, direct calls to the MKL were used.

The Poisson equation on a cube is discretized with $\mathbb{P}_1$ finite elements in double precision with 32-bit integers. This yields a matrix $A$ of dimension one million. Though the numerical values of $A$ are of no interest here, the sparsity pattern is rather common and frequently encountered, e.g., when discretizing partial differential equations. Then, a random symmetric dense matrix $T$ of dimension $b$ is generated, for $b \in \{ 1,3,6 \}$, and the following matrix is evaluated $\mathbb{A} = A \otimes T$. The matrix $\mathbb{A}$ is then assembled into all three formats described above. For block matrices, a block size of $b$ is used. The performance of the \pk{MatProductNumeric} operation is evaluated for tall-and-skinny dense matrices with a varying number of columns $N \in \{1,2,8,16,32,64\}$. For $N=1$, the \pk{MatMult} operation for matrix--vector multiplication is used instead. All results are gathered in~\cref{fig:perfAX}. They can be reproduced using the following mini-app: \url{http://jolivet.perso.enseeiht.fr/MatProduct.c} and any input \pk{MatAIJ} stored in binary format, here using the default name \pk{binaryoutput}.

@@

  % \pgfmathparse{1.3535e+09/8.9584e-01/1000000000.0}\edef\storeresultII{\pgfmathresult}
  % \pgfmathparse{5.4450e+09/4.0734e+00/1000000000.0}\edef\storeresultIII{\pgfmathresult}
  \pgfmathparse{1.3535e+09/1.2072e+00/1000000000.0}\edef\storeresultII{\pgfmathresult}
  \pgfmathparse{5.4450e+09/5.0428e+00/1000000000.0}\edef\storeresultIII{\pgfmathresult}

@@

@@

@@

          \addplot+[nodes near coords style={xshift=0.5cm,yshift=-0.1cm,above}, nodes near coords*={\rotatebox{45}{\tiny\color{black}{\pgfmathfloattofixed{\pgfplotspointmeta}\pgfmathparse{int(round(\pgfmathresult))}\ifnum\pgfmathresult=1\pgfmathprintnumber\storeresultII~GFLOP/s\fi}}},draw=black,fill, area legend,postaction={pattern=crosshatch dots}] table [x expr=\thisrow{N}>7?\thisrow{N}/2:\thisrow{N},y expr=\thisrow{aij-3-F}/\thisrow{aij-3-T}/(\storeresultII*1000000000.0)]{\loadedMatProduct};\label{pgfplots:aij3}
          \addplot+[

@@

@@

        ytick = {1,3,5},

@@

          \addplot+[nodes near coords style={xshift=0.5cm,yshift=-0.1cm,above}, nodes near coords*={\begin{pgfonlayer}{main}\rotatebox{45}{\tiny\color{black}{\pgfmathfloattofixed{\pgfplotspointmeta}\pgfmathparse{int(round(\pgfmathresult))}\ifnum\pgfmathresult=1\contour{white}{\pgfmathprintnumber\storeresultIII}~GFLOP/s\fi}}\end{pgfonlayer}},draw=black,fill, area legend,postaction={pattern=crosshatch dots}] table [x expr=\thisrow{N}>7?\thisrow{N}/2:\thisrow{N},y expr=\thisrow{aij-10-F}/\thisrow{aij-10-T}/(\storeresultIII*1000000000.0)]{\loadedMatProduct};\label{pgfplots:aij10}
          \addplot+[draw=black,fill, area legend,postaction={pattern=crosshatch dots}] table [x expr=\thisrow{N}>7?\thisrow{N}/2:\thisrow{N},y expr=\thisrow{baij-10-F}/\thisrow{baij-10-T}/(\storeresultIII*1000000000.0)]{\loadedMatProduct};\label{pgfplots:baij10}

@@

   \caption{Performance of the matrix--matrix multiplication as implemented in PETSc for three sparse matrix types. As baseline values with $N=1$, performance of \pk{MatAIJ} matrix--vector multiplication is used~\ref{pgfplots:aij1}. Results are averages of 5 consecutives multiplications.\label{fig:perfAX}}

@@

Some conclusions may be drawn:
    \item with a block size of 1, top plot $b=1$ in~\cref{fig:perfAX}, using any type but \pk{MatAIJ} is counterproductive, see~\ref{pgfplots:baij1} and~\ref{pgfplots:sbaij1}, even with the MKL, see~\ref{pgfplots:mklbaijseq1} and~\ref{pgfplots:mklsbaijseq1}. However, with this format, as soon as the number of columns is greater or equal than 8, the efficiency reaches approximately 200\%, see~\ref{pgfplots:aij1}; PETSc implementation seems slightly better than MKL except for the single-column case;
